---
title: "N-glycopeptide Statistical Analysis V.2"
author: "Maarten van Schaik, University of Leeds"
date: "6/28/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Introduction  

This Markdown file contains code used for the a report on the technical variability in N-glycopeptide intensity by Frania J. Zuñiga-Bañuelos. All statistical analysis is done by Maarten van Schaik. Frania and Maarten are PhD candidates in the EU-funded programme [IMforFUTURE](https://imforfuture.eu) (Innovative training in methods for future data).

The report can be read at [GitHub](https://mjgvanschaik.github.io/glyXera_secondment/)

For the statistical part in the report, the following software was used:

`R` version 3.6.3 (2020-02-29) 64-bit
Tableau 2020.2

The following R packages were used:

`Tidyverse` 1.3.0
`readxl` 1.3.1
`writexl` 1.3
`ggplot2` 3.3.0
`data.table` 1.12.8

## Setup

The setup block should always be run first since it loads the required packages and sets the correct location of the data. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
require(readxl)
require(writexl)
require(ggplot2)
require(data.table)
```

# Load data

This block of code will load the 4 sample files and combine them into one main file which will be the basis of all further analyses. 

We only analyse the rows for which `Validate` = `True-positive`, `Glycans` is an observed glycan, and `Quant.level` = `1`. 

```{r}
data_location <- "data/"
fib1 <- read.csv(paste0(data_location, "Fib_1.csv")); fib1$batch <- "1"; fib1$row.number <- 1:nrow(fib1)
fib2 <- read.csv(paste0(data_location, "Fib_2.csv")); fib2$batch <- "2"; fib2$row.number <- 1:nrow(fib2)
fib3 <- read.csv(paste0(data_location, "Fib_3.csv")); fib3$batch <- "3"; fib3$row.number <- 1:nrow(fib3)
fib4 <- read.csv(paste0(data_location, "Fib_4.csv")); fib4$batch <- "4"; fib4$row.number <- 1:nrow(fib4)


fib <- rbind(fib1, fib2, fib3, fib4)

fib <- fib %>%
  filter(Validate=="True-positive",
         Glycans!="",
         Quant.level==1)

fib$batch <- as.factor(fib$batch)

fib$Protein <- as.character(fib$Protein.name)
fib$Protein[grepl("alpha", fib$Protein)] <- "alpha"
fib$Protein[grepl("beta", fib$Protein)] <- "beta"
fib$Protein[grepl("gamma", fib$Protein)] <- "gamma"

fib$Peptide <- droplevels(fib$Sequence..unformatted.)
fib$Peptide <- as.character(fib$Peptide)
fib$Peptide[grepl("DLQSLEDILHQVENK", fib$Peptide)] <- "(VDK)DLQSLEDILHQVENK"
fib$Peptide <- factor(fib$Peptide)

fib$Glycans <- droplevels(fib$Glycans)

fib$XIC.AUC <- as.numeric(as.character(fib$XIC.AUC))
fib$XIC.area.summed <- as.numeric(as.character(fib$XIC.area.summed))
fib$Total.XIC.AUC.Averagine <- as.numeric(as.character(fib$Total.XIC.AUC.Averagine))

table(fib$Peptide)
table(fib$Protein)
table(fib$Glycans)

fib <-  fib %>% 
  mutate(PPG = as.factor(paste0(Protein, ";",
                      Peptide, ";",
                      Glycans)))

rm(data_location, fib1, fib2, fib3, fib4)

write.csv(fib, file="data/Fib_total_processed.csv")

saveRDS(fib, "data/Fib_total_processed.Rds")
```


# Mean, sd and CV statistics for number of glycopeptide observations

The first analyses we did was to look at the number of observations, the number of observations relative to the total number of observations, and a presence/absence indicator. 

```{r}
fib <- readRDS("data/Fib_total_processed.Rds")

# Take data,
fib_summary <- fib %>%
  # Group by PPG-batch combination, do not drop combinations with zeros
  group_by(PPG,
           batch,
           .drop=FALSE) %>%
  # For each combination, return the count and a presence indicator
  summarise(n=n(),
            binary = n > 0) %>%
  # Now group by batch only
  group_by(batch) %>%
  # Calculate the total nr of observations per batch 
  # and the relative counts. 
  mutate(batch_total = sum(n),
         abundance = n/batch_total)

# Now we create a table with summaries per N-glycopeptide (PPG). 
# Take data,
fib_summary2 <- fib_summary %>%
  # Group by PPG,
  group_by(PPG) %>%
  # Calculate mean, sd, CV of relative counts, counts, and binary presences.
  summarise(mean_abundance = mean(abundance),
            sd_abundance = sd(abundance),
            CV_abundance = sd_abundance / mean_abundance,
            mean_count = mean(n),
            sd_count = sd(n),
            CV_count = sd_count / mean_count,
            mean_presence = mean(binary),
            mean_presence = mean(binary),
            sd_presence = sd(binary),
            CV_presence = sd_presence / mean_presence)

# Rename the 'levels' of the PPG variable to be shorter ("PPG#").
fib_summary$PPG_nr <- as.factor(fib_summary$PPG)
fib_summary2$PPG_nr <- as.factor(fib_summary2$PPG)
levels(fib_summary$PPG_nr) <- paste0("PPG-", 1:length(levels(fib_summary$PPG)))
levels(fib_summary2$PPG_nr) <- paste0("PPG-", 1:length(levels(fib_summary2$PPG)))

# Create a "wide" table to summarize our findings on the N-glycopeptide counts. 
fib_summary_count <- right_join(fib_summary %>%
                                  dplyr::select(PPG_nr,
                                                PPG,
                                                batch,
                                                n) %>%
                                  spread(batch,
                                         n),
                                fib_summary2 %>%
                                  dplyr::select(PPG_nr,
                                                # PPG,
                                                mean_count,
                                                sd_count,
                                                CV_count),
                                "PPG_nr")%>%
  separate(col=PPG,
           into=c("protein", "peptide", "glycan"),
           sep=";")

# Create a "wide" table to summarize our findings on the N-glycopeptide abundances.
fib_summary_abundance <- right_join(fib_summary %>%
                                      dplyr::select(PPG_nr,
                                                    PPG,
                                                    batch,
                                                    abundance) %>%
                                      spread(batch,
                                             abundance),
                                    fib_summary2 %>%
                                      dplyr::select(PPG_nr,
                                                    # PPG,
                                                    mean_abundance,
                                                    sd_abundance,
                                                    CV_abundance),
                                    "PPG_nr")%>%
  separate(col=PPG,
           into=c("protein", "peptide", "glycan"),
           sep=";")

# Create a "wide" table to summarize our findings on the N-glycopeptide presences.
fib_summary_presence <- right_join(fib_summary %>%
                                     dplyr::select(PPG_nr,
                                                   PPG,
                                                   batch,
                                                   binary) %>%
                                     spread(batch,
                                            binary),
                                   fib_summary2 %>%
                                     dplyr::select(PPG_nr,
                                                   # PPG,
                                                   mean_presence,
                                                   sd_presence,
                                                   CV_presence),
                                   "PPG_nr") %>%
  separate(col=PPG,
           into=c("protein", "peptide", "glycan"),
           sep=";")

# Reshape the 'wide' tables created above to be 'long' format. 
# This is useful for making plots with ggplot2 and Tableau. 
fib_summary_count_long <- gather(fib_summary_count,
                                 batch,
                                 count,
                                 "1":"4")

fib_summary_abundance_long <- gather(fib_summary_abundance,
                                     batch,
                                     abundance,
                                     "1":"4")

fib_summary_presence_long <- gather(fib_summary_presence,
                                    batch,
                                    presence,
                                    "1":"4")

# Export the long format lists to an Excel file with multiple sheets.
# This is the file we import into Tableau. 
write_xlsx(list("fib_summary_count" = fib_summary_count_long,
                "fib_summary_abundance" = fib_summary_abundance_long,
                "fib_summary_presence" = fib_summary_presence_long),
           "data/summaries/fib_summary_counts.xlsx")

# We can also create plots in R.
# three separate plots
plot(ecdf(fib_summary_count$CV_count),
     xlab="CV",
     ylab="Percentage",
     main="Empirical Cumulative Distribution: CV of P-P-G counts")

plot(ecdf(fib_summary_abundance$CV_abundance),
     xlab="CV",
     ylab="Percentage",
     main="Empirical Cumulative Distribution: CV of P-P-G relative counts")

plot(ecdf(fib_summary_presence$CV_presence),
     xlab="CV",
     ylab="Percentage",
     main="Empirical Cumulative Distribution: CV of P-P-G presence")

# one plot
plot(ecdf(fib_summary_count$CV_count),
     xlab="CV",
     ylab="Percentage",
     main="Empirical Cumulative Distribution: CV of PPGs",
     xlim=c(0, 3),
     col="green")
lines(ecdf(fib_summary_abundance$CV_abundance),
      col="blue")
lines(ecdf(fib_summary_presence$CV_presence),
      col="red")
legend('bottomright',
       legend=c("count","abundance","presence"),  # text in the legend
       col=c("green","blue","red"),  # point colors
       pch=15)  # specify the point type to be a square

# Plot CV as a function of mean abundance.
plot(fib_summary_abundance$CV_abundance~fib_summary_abundance$mean_abundance,
     xlab="PPG mean abundance",
     ylab="PPG CV of abundance",
     main="Mean-CV relationship of PPG relative counts")
```


```{r}
fib_summary3 <- fib %>% 
  mutate(PPG = as.factor(paste0(Protein, ";",
                                Peptide, ";",
                                Glycans))) %>% 
  group_by(PPG,
           # batch,
           .drop=FALSE) %>% 
  summarise(n=n(),            
            mean_auc = mean(XIC.AUC),
            sd_auc = sd(XIC.AUC),
            CV_auc = sd_auc / mean_auc)

# one plot
plot(ecdf(fib_summary3$CV_auc),
     xlab="CV",
     ylab="Percentage",
     main="Empirical Cumulative Distribution: CV of PPGs",
     xlim=c(0, 3),
     col="green")

```

# Mean, sd and CV statistics for the absolute XIC.AUC intensity 

The code below will investigate the absolute "intensity" as defined by integrated (summed) AUC scores of the XIC.

```{r}
fib <- readRDS("data/Fib_total_processed.Rds")

fib_AUC_total <- fib %>% 
  group_by(batch) %>% 
  summarise(XIC.AUC_batch.total = sum(XIC.AUC))

fib_PPG_summary <- fib %>% 
  group_by(PPG,
           batch,
           .drop=FALSE) %>%
  summarise(XIC.AUC_batchPPG.total = sum(XIC.AUC))

fib_summary_AUC <- fib_PPG_summary %>%
  group_by(PPG) %>% 
  summarise(mean_AUC = mean(XIC.AUC_batchPPG.total),
            sd_AUC = sd(XIC.AUC_batchPPG.total),
            CV_AUC = sd_AUC / mean_AUC)

ggplot(data=fib_PPG_summary,
       aes(x=XIC.AUC_batchPPG.total,
           y=PPG)) +
  geom_boxplot()

# ECDF plot of CV
plot(ecdf(fib_summary_AUC$CV_AUC),
          xlab="CV",
          ylab="Percentage",
          main="Empirical Cumulative Distribution: CV of glycopeptide total AUCs")


plot(log2(fib_summary_AUC$sd_AUC) ~ log2(fib_summary_AUC$mean_AUC),
     xlab="Log2(mean(total intensity))",
     ylab="Log2(sd(total intensity))",
     main="sd. versus mean intensity of the 28 glycopeptides in 6 Fibs")

# write_csv(fib_summary_AUC,
#           "Fib_summary_AUC.csv")

```

# Mean, sd and CV statistics for the relative XIC.AUC intensity

Our final analysis focuses on the relative integrated N-glycopeptide intensity as defined by the sum of the Area Under the Curve of the XIC.

```{r}
fib <- readRDS("data/Fib_total_processed.Rds")

# Total AUC per replicate:
# Take the data (validated ions),
fib_AUC_total <- fib %>% 
  # Group by batch (=replicate)
  group_by(batch) %>% 
  # return as summary measure the total AUC of all ions. 
  summarise(XIC.AUC_batch.total = sum(XIC.AUC))

# Summary statistics per PPG
# Take the data,
fib_PPG_summary <- fib %>% 
  # group by PPG and batch combination (also return the ones with 0 observations),
  group_by(PPG,
           batch,
           .drop=FALSE) %>%
  # Calculate the total AUC per combination. 
  summarise(XIC.AUC_batchPPG.total = sum(XIC.AUC)) %>% 
  # Then merge with the table of total AUC per replicate,
  left_join(fib_AUC_total,
             by="batch") %>% 
  # and calculate the PPG AUC relative to the total AUC per replicate. 
  mutate(XIC.AUC_batchPPG.relative = XIC.AUC_batchPPG.total / XIC.AUC_batch.total)

# Create a 'wide' style table of our findings. 
fib_summary_AUC <- left_join(fib_PPG_summary %>% 
                               dplyr::select(PPG, batch, XIC.AUC_batchPPG.relative) %>% 
                               spread(batch, 
                                      XIC.AUC_batchPPG.relative),
                             fib_PPG_summary %>% 
                               group_by(PPG) %>% 
                               summarise(mean_AUC.relative = mean(XIC.AUC_batchPPG.relative),
                                         sd_AUC.relative = sd(XIC.AUC_batchPPG.relative),
                                         CV_AUC.relative = sd_AUC.relative / mean_AUC.relative),
                             "PPG")

# create a 'long' style table of our findings. 
fib_summary_AUC_long <-  
  fib_summary_AUC %>% 
  gather(batch, AUC.relative,
         "1":"4")

# Boxplot for relative intensities
ggplot(data=fib_summary_AUC_long,
       aes(x=AUC.relative,
           y=PPG)) +
  xlab("Relative intensity") +
  ylab("Protein-Glycopeptide") + 
  ggtitle("Relative intensity of glycopeptides") + 
  geom_boxplot()

# Point-plot for CVs of relative intensities
ggplot(data=fib_summary_AUC_long,
       aes(x=CV_AUC.relative,
           y=PPG)) +
  xlab("CV of relative intensity") + 
  ylab("Protein-Glycopeptide") + 
  ggtitle("CV of glycopeptide relative intensity") + 
  geom_point(size=3)

# Boxplot and Point-plot overlaid together.
# I also made this plot in Tableau: it looks better that way. 
# Still, here is the ggplot2 code:
# FYI: how ggplot works is that it will plot the values of the intensities and CV values at the same time,
#  so we must scale the two variables to the desired (same) scale, and then adjust the axes to hide the fact that we did this. 
coeff <- max(fib_summary_AUC_long$CV_AUC.relative) 
# intensitycolour <- "#69b3a2"
intensitycolour <- "black"
CVcolour <- rgb(0.2, 0.6, 0.9, 1)
ggplot(data=fib_summary_AUC_long,
       aes(y=PPG)) + 
  geom_boxplot(aes(x=AUC.relative), color=intensitycolour) +
  geom_point(aes(x=CV_AUC.relative / coeff), size=3, color=CVcolour) +
  scale_x_continuous(
    # Feature of the first axis
    name="Relative intensity",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*coeff,
                        name="CV of relative intensity")) +
  theme(axis.title.x = element_text(color = intensitycolour, size=13),
        axis.title.x.top = element_text(color = CVcolour, size=13)) +
  ggtitle("(CV of) glycopeptide relative intensities")

# ECDF plot of CV
plot(ecdf(fib_summary_AUC$CV_AUC.relative),
          xlab="CV",
          ylab="Percentage",
          main="Empirical Cumulative Distribution: CV of glycopeptide relative intensity")

# log2(sd) as a function of log2(mean AUC)
plot(log2(fib_summary_AUC$sd_AUC.relative) ~ log2(fib_summary_AUC$mean_AUC.relative),
     xlab="Log2(mean(relative intensity))",
     ylab="Log2(sd(relative intensity))",
     main="sd. versus mean intensity of the 28 glycopeptides in 6 Fibs")

# Export the long format lists to an Excel file with multiple sheets.
# This is the file we import into Tableau. 
write_xlsx(list("Fib_summary_AUC_wide" = fib_summary_AUC,
                "Fib_summary_AUC_long" = fib_summary_AUC_long),
           "data/summaries/Fib_summary_AUC.xlsx")
```